---
title: "Excercise 2"
subtitle: "TMA4300 Computer intensive statistical methods Spring 2021"
author: 
- "Johan Fredrik Agerup"
- "Arne Rustad"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

\newcommand{\Var}{\textrm{Var}}
\newcommand{\E}{\textrm{E}}
\newcommand{\Cov}{\textrm{Cov}}

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3)
```


```{r rpackages,eval=TRUE,echo=FALSE}
library(ggplot2)
# library(expm)
# library(Matrix)
library(dplyr)
library(tidyverse)
library(boot)
```

# Problem A

In this problem we analyze a data set of time intervals between successive coal-mining disaster in the UK involving ten or more men killed. The data is for the period March 15th 1851 to March 22nd 1962.

You should note that the first and last records in this variable are the start and end dates, respectively. IS THIS TRUE???????????????????????????????????????????????????????????????????????????????????????????????????????????????

First we try to get an impression of the data set by making a plot with year along the $x$-axis and the cumulative number of disasters along the $y$-axis.


```{r, fig.cap = "\\label{fig:1a_cumplot} Cumulative umber of explosions which resulted in 10 or more fatalities. The time span is from March 15 1851 until March 22 1962."}
df.coal = coal
df.coal$cum.n.explosions = cumsum(rep(1, nrow(coal)))
head(df.coal)
ggplot(df.coal, aes(x = date, y = cum.n.explosions)) + geom_line() + ggtitle("Cumulative number of explosions (15.03.1851-22.3.1962)") + xlab("Year") + ylab("Cumlative number of explosions") + geom_vline(xintercept = c(1890,1930))
?
```

From Figure \ref{fig:1a_cumplot} the rate of accidents appear approximately constant from year 1850 until around 1890. Then the rate of large explosions appear to dampen a bit, perhaps due to better safety rutines, equipment, societal norms and laws or another reasson.
